Reproducible Research - Peer Assessment 1
===============================================================================
Matthew Tanner  
mjtanner@uga.edu


### Loading and preprocessing the data

Loading the data requires the fully qualified path and file name or setting the
working directory to the file containing the data as I do here:

```{r data load, echo = TRUE}
setwd("~/RepData_PeerAssessment1")
activityData <- read.table(unz("activity.zip", "activity.csv"), header=T, sep=",")
str(activityData)
```


### What is the mean total number of steps taken per day?

Here we call the dplyr package to call the 'sum' funciotn on all the levels of 'date'.

```{r dply call, echo = TRUE, message = FALSE}
library(dplyr)
totalStepsByDate <- group_by(activityData, date) %>% summarize(total_steps = sum(steps))
hist(totalStepsByDate$total_steps, main = "Histogram of Daily Total Steps", xlab = "Daily Total Steps", breaks = 10)
````


### What is the average daily activity pattern.

Here we again call on the dplyr package to calculate the means steps for all levels of interval and output the results to a histogram.

```{r daily activity patter, echo = TRUE}
averageStepsPerInterval <- group_by(activityData, interval) %>% summarize(average_steps = mean(steps, na.rm = TRUE))
plot(averageStepsPerInterval$interval,averageStepsPerInterval$average_steps, main = "Average Step by Interval", xlab = "Interval", ylab = "Average Steps", type = "l")
```


The following code segment returns the 5-minute interval with the maximun average number of steps

```{r maximum average step 5-minute interval, echo=TRUE}
averageStepsPerInterval$interval[which.max(averageStepsPerInterval$average_steps)]
```


### Imputing missing data


Here we calculate the total number of missing data:

```{r missing data, echo = TRUE}
sum(is.na(activityData$steps))
```


Our strategy is to replace missing data with the weekday average with respect to 5-minute interval.

```{r impute missing date, echo = TRUE}
# New copy of the data.
newActivityData <- activityData

# Introduce the weekday variable.
newActivityData$day <- weekdays(as.Date(newActivityData$date))

# Call dplyr to calculate the mean for all levels of weekday.
averageStepsPerWeekdayInterval <- group_by(newActivityData, interval, day) %>% summarize(average_steps = mean(steps, na.rm = TRUE))

# Strobe all the data records with missing data and replace the missing data with the weekday averate. 
for (i in 1:nrow(newActivityData)) {
  if(is.na(newActivityData[i,,]$steps)) {
    newActivityData[i,]$steps <- as.integer(round(averageStepsPerWeekdayInterval[which(averageStepsPerWeekdayInterval$day == newActivityData[i,]$day & averageStepsPerWeekdayInterval$interval == newActivityData[i,]$interval),]$average_steps))
  }
}
```


We again call on the dplyr functionality to calculate and displace the total number of steps per date. The histogram profile appears unchanged.

```{r total steps missing data, echo = TRUE}
totalStepsByDate <- group_by(newActivityData, date) %>% summarize(total_steps = sum(steps))

hist(totalStepsByDate$total_steps, main = "Histogram of Daily Total Steps (Missing Data Replaced)", xlab = "Daily Total Steps", breaks = 10)
```


The daily mean and median steps are:

```{r mean and median daily, echo = TRUE}
meanSteps = mean(totalStepsByDate$total_steps)
medianSteps = median(totalStepsByDate$total_steps)
```


#Are there differences iun activity patterns between weekdays and weekdays


We first introduce the new variable "daytype".

```{r dayType, echo = TRUE}
dayType <- ifelse((newActivityData$day == "Saturday" | newActivityData$day == "Sunday"), "weekend", "weekday")

newActivityData$dayType <- as.factor(dayType)
```


We then calculate the mean steps for levels of dayType and plot the results.

```{r mean histogram per dayType, echo = TRUE}
averageStepsPerDayTypeInterval <- group_by(newActivityData, interval, dayType) %>% summarize(average_steps = mean(steps))

xWeekday <- averageStepsPerdayTypeInterval$interval[which(averageStepsPerdayTypeInterval$dayType == "weekday")]
xWeekend <- averageStepsPerdayTypeInterval$interval[which(averageStepsPerdayTypeInterval$dayType == "weekend")]
yWeekday <- averageStepsPerdayTypeInterval$average_steps[which(averageStepsPerdayTypeInterval$dayType == "weekday")]
yWeekend <- averageStepsPerdayTypeInterval$average_steps[which(averageStepsPerdayTypeInterval$dayType == "weekend")]

par(mfrow=c(2,1))
plot(xWeekend,yWeekend, main = "weekend", type = "l", ylab = "Average Steps", xlab = "Interval")
plot(xWeekday,yWeekday, main = "weekday", type = "l", ylab = "Average Steps", xlab = "Interval")
```

Significant differences between the weekday and weekend profile are apparent.  The weekday profile has a significantly higher peek corresponding to the morning hours and greater over variablity over time.